Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-06-03 src/tex/ms.tex
[showyourwork] overleaf sync

This factorization of the joint posterior into single star posteriors $\mathbb{P}(\vec\Lambda|D_j)$ has the advantage that we do not need to specify the exact number of observational data points beforehand. We can simply train our neural density estimator on single star observations and then at evaluation time, we simply combine as many posterior estimates as we have observational data.

@@

With this it is straight forward to evaluate eq.~\ref{eq:posterior} analytically. In fact, under our assumption the combined posterior is a product between the prior and the product of multiple Gaussians for the individual star posteriors. The latter product is also a Gaussian with mean $\mathbf{\mu'}$ and variance $\mathbf{\sigma'^2}$:
\begin{align}
    \mathbf{\mu'} &= \frac{\sum_{i=1}^{N_{stars}} \frac{\mu_i}{\sigma_i^2}}{\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}} \\
\mathbf{\sigma'}^2 &= \frac1 {\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}}
\end{align}

Further, in our case the prior for the galactic parameters $\Lambda$ is Gaussian as well. Therefore the resulting factorized posterior from eq.~\ref{eq:posterior} is again a Gaussian and can be expressed with mean $\mathbf{\mu}$ and variance $\mathbf{\sigma}$ as:
\begin{align}
\mathbf{\mu} &= \frac{\frac{\mu'}{\sigma'^2}-\frac{(1-N)\mu_ {prior}}{\sigma_ {prior}^2}}{\frac1{\sigma'^2}-\frac{(1-N)}{\sigma_ {prior}^2}} \\
\mathbf{\sigma}^2 &= \frac1 {\frac1{\sigma'^2}-\frac{(1-N)}{\sigma_ {prior}^2}}
\end{align}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/corner_plot_singlestar.pdf}
    \vspace{-.5cm}
    \caption{Corner plot of the posteriors for all six parameters for a single star from the validation set. The gray contours show a kde-estimate of the posterior from our SBI inference and the red dot and lines show the ground truth parameter values. Gray histograms on the diagonal show a kde estimate of the marginals.}
    \label{fig:corner_plot}
    \script{plot_sbc.py}
\end{figure*}

@@

\begin{figure}
%\vspace{-1.3cm}
    \centering
    \includegraphics[width=\columnwidth]{figures/CHEMPY_TNG_yields.png}
    \vspace{-.25cm}
    \caption{Joint posterior for global galactic parameters $\alpha_{\rm IMF}$ and $\log_{10}(\rm N_{Ia})$ of $100$ stars. The ground truth value is shown by the red star, the posterior mean of the SBI inference is shown with the black dot and white ellipses show the $1-3\sigma$ contours of our inference. The gray ellipses show the results for HMC inference performed by \citet{Philcox_2019}. See Sec.~\ref{subsec:chempy_tng} for a full description.}
    \label{fig:CHEMPY_TNG_sbi} 
    \script{chempy_tng_inference.py}
\end{figure}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/CHEMPY_alternative_yields_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_N_star_analysis} but for the mock data created with a different yield set than the training data. See Sec.~\ref{subsec:mocks_wrong_yield} for a full description.}
    \label{fig:CHEMPY_alt_N_star_analysis}
    \script{chempy_alt_yield_inference.py}
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/TNG_sim_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_N_star_analysis} but for the mock data taken from an IllustrisTNG Milky Way-like galaxy. See Sec.~\ref{subsec:tng_sim} for a full description.}
    \label{fig:TNG_N_star_analysis}
    \script{tng_inference.py}
\end{figure*}

@@

\section{\resub{Comparison between HMC and SBI}}
\label{sec:hmc-sbi}
\resub{put a new figure and some text...}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/hmc_corner_plot.pdf}
    \vspace{-.5cm}
    \caption{Caption}
    \label{fig:hmc}
    \script{hmc.py}
\end{figure}

>>>>

This factorization of the \resub{marginal} joint posterior \resub{of the global parameters} into single star posteriors $\mathbb{P}(\vec\Lambda|D_j)$ has the advantage that we do not need to specify the exact number of observational data points beforehand. We can simply train our neural density estimator on single star observations and then at evaluation time, we simply combine as many posterior estimates as we have observational data.

@@

With this it is straight forward to evaluate eq.~\ref{eq:posterior} analytically. In fact, under our assumption the combined \resub{factorized posterior from eq.~\ref{eq:posterior} is again a Gaussian whose mean and covariance can be calculated from the means and covariances of the individual prior/posterior Gaussians as follows:}
\resub{
\begin{align}
    \mathbf{\Sigma}_{\rm Posterior}^{-1} &= (N-1) \mathbf{\Sigma}_{\rm Prior}^{-1} + \sum_i^N \mathbf{\Sigma}_i^{-1} \\
\mathbf{\mu}_{\rm Posterior} &= \mathbf{\Sigma}_{\rm Posterior} \Big( (N-1) \mathbf{\Sigma}_{\rm Prior}^{-1} \mathbf{\mu}_{\rm Prior} + \sum_i^N \mathbf{\Sigma}_i^{-1} \mathbf{\mu}_i \Big)
\end{align}
}

\resub{where we have assumed that the prior for the global parameters has no  covariance such that $\Sigma_\rm{prior}$ is diagonal.}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/corner_plot_comparison_singlestar_17.pdf}
    \vspace{-.5cm}
    \caption{Corner plot of the posteriors for all six parameters for a single star from the validation set. The blue contours show a kde-estimate of the posterior from our SBI inference and the red dot and lines show the ground truth parameter values. The orange contours show results from a MCMC run on the same data. Blue/orange histograms on the diagonal show a kde estimate of the marginals.}
    \label{fig:corner_plot}
    \script{emcee_sampling.py}
\end{figure*}

@@

\begin{figure}
%\vspace{-1.3cm}
    \centering
    \includegraphics[width=\columnwidth]{figures/CHEMPY_TNG_yields.png}
    \vspace{-.25cm}
    \caption{Joint posterior for global galactic parameters $\alpha_{\rm IMF}$ and $\log_{10}(\rm N_{Ia})$ of $100$ stars. The ground truth value is shown by the red star, the posterior mean of the SBI inference is shown with the blue dot and blue ellipses show the $1-3\sigma$ contours of our inference.  See Sec.~\ref{subsec:chempy_tng} for a full description.}
    \label{fig:CHEMPY_TNG_sbi} 
    \script{chempy_tng_inference.py}
\end{figure}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/CHEMPY_alternative_yields_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_N_star_analysis} but for the mock data created with a different yield set than the training data. See Sec.~\ref{subsec:mocks_wrong_yield} for a full description.}
    \label{fig:CHEMPY_alt_N_star_analysis}
    \script{chempy_alt_yield_inference.py}
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/TNG_sim_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_N_star_analysis} but for the mock data taken from an IllustrisTNG Milky Way-like galaxy. See Sec.~\ref{subsec:tng_sim} for a full description.}
    \label{fig:TNG_N_star_analysis}
    \script{tng_inference.py}
\end{figure*}

@@

\section{\resub{Comparison between MCMC and SBI}}
\label{sec:hmc-sbi}
\resub{Figure~\ref{fig:hmc} shows a comparison of the marginal joint posterior of the global parameters for 200 stars between MCMC (orange) and our SBI inference (blue). As already shown for a single star in Fig.~\ref{fig:corner_plot} our SBI results converge nicely to the MCMC results also when combining individual inferences over 200 stars.}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/mcmc_results.pdf}
    \vspace{-.5cm}
    \caption{Marginal joint posterior of the global parameters for both SBI (blue) and MCMC (orange) for 200 stars. Our SBI inference converges nicely to the result of the MCMC inference.}
    \label{fig:hmc}
    \script{emcee_sampling.py}
\end{figure}
